Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)

### 2017-02-15 csf-2017/paper.tex
reformatted subsections of discussion section

\section{Discussion}
\label{sec:discussion}
\subsection{Calling convention}
\textbf{Formulating control flow correctness} While we claim that our calling convention enforces control-flow
correctness, we do not prove a general theorem that shows this, it is not clear what such a theorem should state. Formulations in terms of a control-flow graph, like the one by
\cite{abadi_control-flow_2005}, do not take into account temporal properties,
like the well-bracketedness that Example~\texttt{\footnotesize{g1}} relies on.
In fact, our examples show that our logical relation imply a form of
control-flow correctness that is stronger than such formulations, although this
is not made very explicit. As future work, we consider looking at a more
explicit and useful way to formalise well-bracketed control flow. The idea would
be to define a variant of our capability machine with built-in call and return
instructions and well-bracketed control flow built-in to the operational
semantics, and then prove that compiling such code to our machine using our
calling convention is fully abstract~\cite{abadi_protection_1998}.

\textbf{Performance and the requirement for stack clearing} 
The additional security measures of the calling convention described in
Section~\ref{sec:stack-and-return-pointer} impose an overhead compared to
a calling convention that does not provide security. However, the security measures generally consist of a
few checks or register clearings on every boundary crossing between trusted code
and adversary, which should impose an acceptable performance overhead. The only
exception is the requirements for stack clearing which we occur in two
situations: when returning to the adversary and when invoking an adversary
callback. As we have explained, it is not sufficient to clear just the part of
the stack that we have used ourselves, but we need to clear all of the stack that we are not using ourselves. 
In other words on every boundary cross between trusted code and adversary code, a potentially large region of memory must be erased.
We believe this is a common requirement for typical usage scenarios of local capabilities and
capability machines like CHERI should consider to provide special support for
this requirement, in the form of a highly-optimized instruction for erasing
a large block of memory.

\textbf{Modularity} It is important that our calling convention is modular.
This means that the requirements we have on callbacks, return
pointers etc. provided to trusted code by an adversary are also satisfied by
callbacks, return pointers etc. that we construct ourselves. For example, our
return pointers are local capabilities because they must point to
memory where we can store the old stack pointer, but the adversary's return
pointers are also allowed to be local. Adversary callbacks on the other hand are
required to be global but the callbacks we construct ourselves are also
allocated on the heap and global. In other words, we do not assume that our code
has some sort of special status and it is possible for the adversary to protect
themselves against us the same way we protect against them.

\textbf{Single orthogonal closure} The definitions of $\stder$ and $\stdvr$ in
Figure~\ref{fig:logrel} apply a single orthogonal closure, a new variant of an
existing pattern called biorthogonality. Biorthogonality is a pattern for
defining logical relations~\citep{krivine_classical_1994,pitts_operational_1998}
in terms of an observation relation of safe configurations (like we do). The
idea is to define safe evaluation contexts as the set of contexts that produce
safe observations when plugging safe values and define safe terms as the set of
terms that can be plugged into safe evaluation contexts to produce safe values.
This is an alternative to more direct definitions where safe terms are defined
as terms that evaluate to safe values. An advantage of biorthogonality is that
it scales better to languages with control effects like call/cc. Our definitions
can be seen as a variant of biorthogonality, where we take only a single
orthogonal closure: we do not define safe evaluation contexts but immediately
define safe terms as those that produce safe observations when plugged with safe
values. This is natural because we model arbitrary assembly code that does not
necessarily respect a particular calling convention. As a result, return
pointers are in principle values like all others and there is no reason to treat
them specially in the logical relation.

@@

\textbf{Public/private future worlds} A novel aspect of our logical relation is
how we model the temporary nature of local capabilities using public/private
future worlds. The main insight is that the temporary, revokable nature of these
capabilities can be seen as a generalisation of the syntactically-enforced
unstoreable status of evaluation contexts in lambda calculi without control
effects (of which well-bracketed control flow is a consequence). To reason about
code that relies on this (particularly, the original awkward example),
\citet{Dreyer:jfp12} (DNB) formally capture the special status of evaluation
contexts using Kripke worlds with public and private future world relations.
Essentially, they allow relatedness of evaluation contexts to be monotone with
respect to a weaker future world relation (public) than relatedness of values,
formalising the idea that it is safe to make temporary internal state
modifications (private world transititions, that invalidate the continuation,
but not other values) while an expression is performing internal steps, as long
as the code returns to a stable state (i.e. transitions to a public future world
of the original) before returning (i.e. invoking the continuation). We
generalise this idea to reason about local capabilities: validity of local
capabilities is allowed to be monotone with respect to a weaker future-world
relation than other values, which we can exploit to distinguish between state
changes that are always safe (public future worlds) and changes that are only
valid if we clear all local capabilities (private future worlds). Our future
world relations are similar to \citet{Dreyer:jfp12}'s (for example, our proof of
the awkward example uses exactly the same state transition system),
but they turn up in an entirely different place in the logical relation: rather
than using public future worlds for the special syntactic category of evaluation
contexts, they are used in the value relation depending on the
locality of the capability at hand. Additionally, our worlds are a bit more
complex because, to allow local memory capabilities and write-local
capabilities, they can contain (revokable) temporary regions that are only
monotonous w.r.t. public future worlds, while DNB's worlds are entirely
permanent.

\textbf{Local capabilities in high-level languages} It is worth mentioning that
local capabilities are quite similar to a feature proposed for the high-level
language Scala: \citet{osvald_gentrification_2016}'s second-class or local
values. They are a kind of values that can be provided to other code for
immediate use without allowing it to be stored in a closure or reference for
later use. We believe reasoning about such values will require techniques similar 
to the ones we have used for local capabilities.

>>>>

\section{Discussion}
\label{sec:discussion}
\noindent\textbf{Calling convention}\\
\emph{Formulating control flow correctness} While we claim that our calling convention enforces control-flow
correctness, we do not prove a general theorem that shows this, it is not clear what such a theorem should state. Formulations in terms of a control-flow graph, like the one by
\cite{abadi_control-flow_2005}, do not take into account temporal properties,
like the well-bracketedness that Example~\texttt{\footnotesize{g1}} relies on.
In fact, our examples show that our logical relation imply a form of
control-flow correctness that is stronger than such formulations, although this
is not made very explicit. As future work, we consider looking at a more
explicit and useful way to formalise well-bracketed control flow. The idea would
be to define a variant of our capability machine with built-in call and return
instructions and well-bracketed control flow built-in to the operational
semantics, and then prove that compiling such code to our machine using our
calling convention is fully abstract~\cite{abadi_protection_1998}.

\emph{Performance and the requirement for stack clearing} 
The additional security measures of the calling convention described in
Section~\ref{sec:stack-and-return-pointer} impose an overhead compared to
a calling convention that does not provide security. However, the security measures generally consist of a
few checks or register clearings on every boundary crossing between trusted code
and adversary, which should impose an acceptable performance overhead. The only
exception is the requirements for stack clearing which we occur in two
situations: when returning to the adversary and when invoking an adversary
callback. As we have explained, it is not sufficient to clear just the part of
the stack that we have used ourselves, but we need to clear all of the stack that we are not using ourselves. 
In other words on every boundary cross between trusted code and adversary code, a potentially large region of memory must be erased.
We believe this is a common requirement for typical usage scenarios of local capabilities and
capability machines like CHERI should consider to provide special support for
this requirement, in the form of a highly-optimized instruction for erasing
a large block of memory.

\emph{Modularity} It is important that our calling convention is modular.
This means that the requirements we have on callbacks, return
pointers etc. provided to trusted code by an adversary are also satisfied by
callbacks, return pointers etc. that we construct ourselves. For example, our
return pointers are local capabilities because they must point to
memory where we can store the old stack pointer, but the adversary's return
pointers are also allowed to be local. Adversary callbacks on the other hand are
required to be global but the callbacks we construct ourselves are also
allocated on the heap and global. In other words, we do not assume that our code
has some sort of special status and it is possible for the adversary to protect
themselves against us the same way we protect against them.
\\\\
\noindent\textbf{Logical relation}\\
\emph{Single orthogonal closure} The definitions of $\stder$ and $\stdvr$ in
Figure~\ref{fig:logrel} apply a single orthogonal closure, a new variant of an
existing pattern called biorthogonality. Biorthogonality is a pattern for
defining logical relations~\citep{krivine_classical_1994,pitts_operational_1998}
in terms of an observation relation of safe configurations (like we do). The
idea is to define safe evaluation contexts as the set of contexts that produce
safe observations when plugging safe values and define safe terms as the set of
terms that can be plugged into safe evaluation contexts to produce safe values.
This is an alternative to more direct definitions where safe terms are defined
as terms that evaluate to safe values. An advantage of biorthogonality is that
it scales better to languages with control effects like call/cc. Our definitions
can be seen as a variant of biorthogonality, where we take only a single
orthogonal closure: we do not define safe evaluation contexts but immediately
define safe terms as those that produce safe observations when plugged with safe
values. This is natural because we model arbitrary assembly code that does not
necessarily respect a particular calling convention. As a result, return
pointers are in principle values like all others and there is no reason to treat
them specially in the logical relation.

@@

\emph{Public/private future worlds} A novel aspect of our logical relation is
how we model the temporary nature of local capabilities using public/private
future worlds. The main insight is that the temporary, revokable nature of these
capabilities can be seen as a generalisation of the syntactically-enforced
unstoreable status of evaluation contexts in lambda calculi without control
effects (of which well-bracketed control flow is a consequence). To reason about
code that relies on this (particularly, the original awkward example),
\citet{Dreyer:jfp12} (DNB) formally capture the special status of evaluation
contexts using Kripke worlds with public and private future world relations.
Essentially, they allow relatedness of evaluation contexts to be monotone with
respect to a weaker future world relation (public) than relatedness of values,
formalising the idea that it is safe to make temporary internal state
modifications (private world transititions, that invalidate the continuation,
but not other values) while an expression is performing internal steps, as long
as the code returns to a stable state (i.e. transitions to a public future world
of the original) before returning (i.e. invoking the continuation). We
generalise this idea to reason about local capabilities: validity of local
capabilities is allowed to be monotone with respect to a weaker future-world
relation than other values, which we can exploit to distinguish between state
changes that are always safe (public future worlds) and changes that are only
valid if we clear all local capabilities (private future worlds). Our future
world relations are similar to \citet{Dreyer:jfp12}'s (for example, our proof of
the awkward example uses exactly the same state transition system),
but they turn up in an entirely different place in the logical relation: rather
than using public future worlds for the special syntactic category of evaluation
contexts, they are used in the value relation depending on the
locality of the capability at hand. Additionally, our worlds are a bit more
complex because, to allow local memory capabilities and write-local
capabilities, they can contain (revokable) temporary regions that are only
monotonous w.r.t. public future worlds, while DNB's worlds are entirely
permanent.

\emph{Local capabilities in high-level languages} It is worth mentioning that
local capabilities are quite similar to a feature proposed for the high-level
language Scala: \citet{osvald_gentrification_2016}'s second-class or local
values. They are a kind of values that can be provided to other code for
immediate use without allowing it to be stored in a closure or reference for
later use. We believe reasoning about such values will require techniques similar 
to the ones we have used for local capabilities.
